NURM: a tool to locate numerical differences in pipelines

### 2019-09-11 paper.tex
add 'Experiment' section and PFS resutls

Recently, scientists began to realize that the results of many 
scientific experiments were not reproducible. According to the previous 
studies, variety of computational infrastructures including workstation 
types, parallelization methods, operating systems, and analysis 
packages are known to influence reproducibility~\cite{Gronenschild2012, 
diethelm2012limits, Glatard2015, bowring2019exploring}.

In particular, the effect of operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} specially on some of the main 
neuroimaging pipelines including FSL, CIVET and FreeSurfer pipelines. 
The binary differences are measured by the checksum method between 
results of the same analyses on different operating systems, which 
indicate a significant disparity.

This irreproducibility issues are reported as the result of the 
creation, propagation, and amplification of small numerical 
differences~\cite{Gronenschild2012, diethelm2012limits, Glatard2015, 
bowring2019exploring}. In this case, the analysis pipelines are said to 
be numerically unstable. Numerical instability is a characteristic of 
the pipelines which amplify small numerical differences and then hamper 
the reproducibility of the analyses depending on the length of the 
pipeline and magnitude of the differences. In many cases, numerical 
instability is an important issue for reproducibility.

There are different solutions to make reproducible analysis including 
containerization techniques that encapsulate software/hardware 
dependencies, provenance capturing tools, and version control systems. 
However, a comprehensive solution requires to fix the numerical 
instabilities. For this purpose, we introduce a Numerical 
Reproducibility Measurement (NURM) tool to identify the processes in 
pipeline that create numerical differences across several runs. In this 
paper, the execution results of pipelines across different operating 
systems are considered to find out the cause of the irreproducibility 
using the interposition techniques.

@@

Figure~\ref{fig:overview} shows an overview of our method. For each tested
condition, Condition 1 and Condition 2, the pipeline is containerized with
Docker, and its interface is described with
Boutiques~\cite{glatard2017boutiques}. \reprozip~\cite{Chirigati2016} is
added to Condition 1 for provenance tracking. Our tool starts by
identifying temporary files and ``multi-write" files, i.e., files written
by more than one process (Figure~\ref{fig:overview-capturing}). To do that,
the tool first executes the pipeline in Condition 1, to obtain a process
graph (Figure~\ref{fig:overview-capturing} - (1)) and a list of result files.
The list of processes that create intermediary or multi-write
files is then generated (Figure~\ref{fig:overview-capturing} - (2)). 
Using this list, the tool modifies Docker image in Condition 1 and
exectutes the pipeline again through it, which turns to produce results
including temporary and overwritten files
(Figure~\ref{fig:overview-capturing} - (3,4)). In the second step, the
processes are labeled from the differences found in their outputs in the
two tested conditions (Figure~\ref{fig:overview-labelling}). To this order,
the tool builds docker image for Condition 2 by modifying pipeline
processes (Figure~\ref{fig:overview-labelling} - (1)) that recognised from 
the process graph. Then, the modified docker
image is executed in Condition 2 and each process is labeled right after
the running by comparing checksum of the results obtained in Condition 1
and output files of the process in Condition 2
(Figure~\ref{fig:overview-labelling} - (2)). We then label processes in 2
categories depending whether they
create differences (represented in red) or not (green). 
The tool replaces output file with the same file obtained in Condition 1 
if process creates differences (Figure~~\ref{fig:overview-labelling} - (3)).
Process labelling
is done through hooks which are defined during the docker image modifying
procedure, and incremental updates in Condition 2, until the end of the
pipeline execution.

@@

The label of processes that read or 
write temporary files is undetermined because these files are deleted during 
the execution. To address this issue, we replace in Condition 1 every process P that 
writes temporary files with a modification of P that first calls P and then
saves all its output files to a read-only directory. This replacement is
done by modifying the PATH environment variable to point to a directory
containing a modified version of the executable called by the process. The
modified executable will run the original executable and then save its
output files (see Listing~\ref{listing:modify-script}). 
This 
solution does not cover the temporary files that are removed by P 
itself, but this is not a problem since these files do not play any role in 
the subsequent steps of the pipeline.

\begin{listing}
  \begin{minted}[frame=single,
    framesep=3mm,
    linenos=false,
    xleftmargin=0pt,
    tabsize=4]{python}
  #!/usr/bin/env python
  import os.path as op
  import shutil

  # Run the original executable using the same arguments
  executed_file = sys.argv[0]
  exec_argv = op.join('original_scripts', executed_file)
  exec_argv = exec_argv + " ".join(sys.argv[1:])
  subprocess.Popen(exec_argv, shell=True).communicate()

@@

The variety of computing infrastructures, particularly operating system, are 
known to influence the reproducibility of computational pipelines because of the
creation of small numerical differences. 
These differences mainly correspond to the mathematical functions implemented 
in different operating system libraries.
For instance, changing the mathematical functions like \emph{expf()} and 
\emph{cosf()} which manipulate the precision of floating-point representations, 
between \emph{glibc} libraries in different operating systems can produce 
small numerical differences.
Furthermore, a similar issue is expected for any operating system which is 
based on \emph{glibc}, the GNU C library.

Indeed, the propagation and amplification of these tiny perturbations by analysis 
pipelines may cause reproducibility issues. We focus on the differences 
created by different version of operating systems in computational pipelines. 
In particular, NURM-tool is tested on the neuroimaging applications which are 
predominantly using mathematical libraries.

@@

The Human Connectome Project (\href{https://www.humanconnectome.org}{HCP}) 
is a significant effort that will acquire and analyze brain connectivity 
data from 1200 healthy adults.
It enables the neuroscience 
research community to discover relationships between brain circuits and 
indivisual behaviors. This helps to understand a wide range of brain disorders.
The HCP project provides (1) database services (ConnectomeDB) for storing and 
sharing primary and processed data freely, and (2) data analysis pipelines that 
are available under an open source license.

The HCP developed a set of pre-processing pipelines to process structural,
functional and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, particularly PreFreeSurfer
and FreeSurfer. A detailed description of the analyses done in the
pipelines is available in~\cite{glasser2013}.
% IF NEEDED, ADD MORE DETAILS ABOUT WHAT THE PIPELINES DO
% The structural pre-processing pipelines consists of PreFreeSurfer, 
% FreeSurfer and PostFreeSurfer. In this paper, 
% we analyzed PreFreeSurfer pipeline which consist of various steps 
% including correction of MR 
% gradient nonlinearity distortions, align the T1w and T2w images using 
% FSL FLIRT, Align native space to MNI template using ACPC and FSL FLIRT, 
% brain extraction usin FSL FLIRT and FNIRT to MNI template, register T2w 
% to T1w using FLIRT's BBR, perform a bias field correction, and register 
% the subject's native structural space to MNI space.
In this paper, we randomly selected 100 unprocessed subjects from the HCP data release S500.
We first quantify the effect of Linux version on
these pipelines. Then we identify the processes in the
pipelines that are responsible for such effects using NURM-tool.

%\begin{figure}
%\centering
%  \includegraphics[width=\columnwidth]{images/overview.png}
%  \caption{An overview of the proposed technique.}
%  \label{fig:overview}
%\end{figure}

%\begin{table}
%\centering
%\begin{threeparttable}
%\caption{An overview of the HCP data (Humman Connectome DB).}
%
%\begin{tabular}{@{}llllll@{}}
%\toprule
%Subject & Release & Acquisition & Gender & Age      & Subj\_Type   \\ \midrule
%103515  & Q1      & Q02         & F      & 26-30    & type1         \\
%105216  & Q3      & Q03         & M      & 26-30    & type4         \\
%103414  & Q2      & Q02         & F      & 22-25    & type1         \\
%125525  & Q1      & Q01         & F      & 31-35    & type1         \\
%142828  & Q1      & Q01         & M      & 31-35    & type1         \\
%129533  & Q3      & Q04         & F      & 31-35    & type1         \\
%103818  & Q1      & Q01         & F      & 31-35    & type3         \\
%133928  & Q2      & Q03         & M      & 26-30    & type3         \\
%148032  & Q3      & Q03         & F      & 31-35    & type2         \\
%139637  & Q2      & Q03         & F      & 31-35    & type1         \\ \bottomrule
%\end{tabular}
%\begin{tablenotes}
%      \small
%      \item *Subj\_Type refers to the tree topology type of the subjects.
%\end{tablenotes}
%\end{threeparttable}
%\label{table:data}
%\end{table}

@@

We executed the pipelines using Docker containers to simplify the 
deployment of different operating system versions on execution 
platforms. The Docker images were built for the HCP pre-processing 
pipelines v3.19.0 (PreFreeSurfer and FreeSurfer) in 
CentOS 6.8 and CentOS 7.2. Docker images are available on 
\href{https://hub.docker.com/r/bigdatalabteam/hcp-prefreesurfer/}{DockerHub}
for reuse.

Two types of errors can occur in the subjects due to the 
errors in the operating systems. One is inter-OS error caused by the 
operating system library updates and the other type, intra-OS errors 
occurs as a result of the pseudo-random processes used in the 
pipelines. 
We have initialized random number generators using a seed state 
to avoid intra-OS errors in our experiments. 
Therefore, NURM-tool is able to identify the process in the pipelines 
that create differences caused by the inter-OS errors.

@@

In the previous study~\cite{Scaria2017}, we showed that the HCP pre-processing 
pipelines~\cite{glasser2013} were 
sensitive to operating system variation so that Figure 
\ref{fig:tissue_class} illustrates the binarized differences after brain 
tissue segmentation process in FreeSurfer. In addition, 
Figure~\ref{fig:fnirt_result} shows the same differences occurred in 
non-linear registration process between CentOS6 and CentOS7 in PreFreeSurfer.

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_classification.png} 
  \caption{Binarized differences between brain segmentation results from 
           FreeSurfer, subject 105216 (CentOS6 vs. CentOS7)~\cite{Scaria2017}.} 
  \label{fig:tissue_class}
\end{figure}

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/fnirt_result.png} 
  \caption{Absolute differences between FNIRT results from PreFreeSurfer 
  (T2w\_acpc\_to\_MNI\_nonli.nii.gz), subject 104820 (CentOS6 vs. 
  CentOS7)~\cite{Scaria2017}. } 
  \label{fig:fnirt_result}
\end{figure}

@@

We clustered subjects into different groups using the NURM-tool, 
after that PreFreeSurfer pipeline has executed on input data.
Figure~\ref{fig:subj-clusters} shows the result of data clustering. 
Among our dataset, we identified 4 types of subjects with 
different topology of \emph{process trees}. In addition, we verified that 
the \emph{process trees} were identical for all subjects of the same type in 
different versions of CentOS operating system.

We identified processes that introduce differences in PreFreeSurfer 
pipeline. 
The processes that introduce differences in PreFreeSurfer along with the 
number of occurrences are depicted in Figure~\ref{fig:pfs_chart} 
including linear registration with “\emph{FLIRT}” (in 
ACPC Alignment, BrainExtraction, DistortionCorrection, 
AtlasRegistration), non-linear registration with “\emph{FNIRT}” (in 
BrainExtraction and AltasRegistration), image warping with 
“\emph{new\_invwarp}” (in BrainExtraction and AtlasRegistration) and 
“\emph{applywarp}” (in DistortionCorrection, BrainExtraction, AtlasRegistration).  
In addition, differences were observed in image mean and standard-deviation 
computations with “\emph{fslstats}” (in BiasFieldCorrection), and in 
masked image extrapolation with “\emph{fslmaths}” (in BiasFieldCorrection). 
Besides, transformation format conversion with 
“\emph{convertwarp}” (in DistortionCorrection) was able to create differences.
It can be seen that \emph{fslmaths}, \emph{applywarp}, and \emph{flirt} 
are the most frequent processes that create differences in the 
PreFreeSurfer pipeline.
\note{add: \emph{wb\_command}}

Furthermore, we determined the frequency of each process that creates differences in 
each pipeline element separately in Figure~\ref{fig:pfs_freq}. It is clear that 
DistortionCorrection element in PreFreeSurfer has the most calls of the processes that 
create differences.
% Among the 117 data files produced by PreFreeSurfer, 21 did 
% not have any error for any subject, 92 had errors for all subjects and 
% 4 had errors for 3 subjects only.

% Figure \ref{fig:complete-graph} shows the annotated provenance graph 
% of the PreFreeSurfer pipeline executed on CentOS6 and CentOS7. Each 
% node in the graph represent an executed process in the pipeline. 
% Processes that created errors are shown in red, processes that removed 
% errors are in blue, and other processes are in green.  Squares denote 
% processes for which the labelling is uncertain, due to temporary 
% files that were removed during the execution. Black edges link 
% sub-processes to their parents while dashed edges denote file 
% dependencies between processes (green edges: files with no errors; red 
% edges: files with errors; yellow edges: temporary files).

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/pfs_barchart.png} \caption{Number 
    of occurrences of processes that create differences in PreFreeSurfer 
    among 100 subjects.} 
    \label{fig:pfs_chart}
  \end{figure}

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/pfs_frequency.png}
  \caption{Colorized processes along with the pipeline elements that introduce 
           differences. The range between red and green colors show the most and 
           least frequency of processes respectively.}
  \label{fig:pfs_freq}
\end{figure}

\subsubsection{FreeSurfer pipeline analysis} 
Result of FreeSurfer\dots

@@

\note{Pipeline amplify small numerical differences because they are numerically 
unstable. Furthermore, math libraries evolve over time, leading to 
different numerical errors. we listed some of the irreproducibility 
causes of the pipelines as we mentioned in the previous section 
overally along with exact command arguments.

@@

The results still suffer from small perturbations literally because of 
the fact that pipeline are not numerically stable. The preferred 
solution is to detect and fix numerical instability of the pipeline 
instead of masking the problem. These processes need to be reviewed to 
understand and correct the cause of instabilities.

It should be noted that processes that remove differrences, read files with differences and 
write files without differences, are not recognised 
by this method.
}

@@

\begin{figure*}
\centering
  \includegraphics[width=.9\textwidth]{images/graph}
  \caption{A complete process graph from the PreFreeSurfer pipeline.
Full-resolution image available at \url{https://drive.google.com/open?id=174yyn8SuVOUcK5aRVw0bagjDanLD0FLt}.}
  \label{fig:complete-graph}
\end{figure*}

\begin{figure*}
\centering
  \includegraphics[width=.8\textwidth]{images/hclusters}
  \caption{Different data types clustered among 100 subjects.}
  \label{fig:subj-clusters}
\end{figure*}

>>>>

\note{Recently, scientists began to realize that the results of many 
scientific experiments were not reproducible.

In particular, validation of the reproducibility has been widely investigated 
in the field of neuroimaging in which uses optimized 
processing methods with the purpose of functional and/or structural 
assessments of the human brain.

According to the previous 
studies, variety of computational infrastructures including workstation 
types, parallelization methods, operating systems, and analysis 
packages are known to influence reproducibility~\cite{Gronenschild2012, 
diethelm2012limits, Glatard2015, bowring2019exploring}.

In particular, the effect of the operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} specially on some of the main 
neuroimaging pipelines including FSL, CIVET and FreeSurfer pipelines. 
The binary differences are measured by the checksum method between 
results of the same analyses on different operating systems, which 
indicate a significant disparity. 
}

These irreproducibility issues are reported as the result of the 
creation, propagation, and amplification of small numerical 
differences~\cite{Gronenschild2012, diethelm2012limits, Glatard2015, 
bowring2019exploring}. In this case, the analysis pipelines are said to 
be numerically unstable. Numerical instability is a characteristic of 
the pipelines which amplify small numerical differences and then hamper 
the reproducibility of the analyses depending on the length of the 
pipeline and magnitude of the differences. In many cases, numerical 
instability is an important issue for reproducibility.

There are different solutions to make reproducible analysis including 
containerization techniques that encapsulate software/hardware 
dependencies, provenance capturing tools, and version control systems. 
However, a comprehensive solution requires to fix the numerical 
instabilities. For this purpose, we introduce a Numerical 
Reproducibility Measurement (NURM) tool to identify the processes in 
the pipeline that create numerical differences across several runs. In this 
paper, the execution results of pipelines across different operating 
systems are considered to find out the cause of the irreproducibility 
using interposition techniques.

@@

Figure~\ref{fig:overview} shows an overview of our method. For each tested
condition, Condition 1 and Condition 2, the pipeline is containerized with
Docker, and its interface is described with
Boutiques~\cite{glatard2017boutiques}. \reprozip~\cite{Chirigati2016} is
added to Condition 1 for provenance tracking. Our tool starts by
identifying temporary files and ``multi-write" files, i.e., files written
by more than one process (Figure~\ref{fig:overview-capturing}). To do that,
the tool first executes the pipeline in Condition 1, to obtain a process
graph (Figure~\ref{fig:overview-capturing} - (1)) and a list of result files.
The list of processes that create intermediary or multi-write
files is then generated (Figure~\ref{fig:overview-capturing} - (2)). 
Using this list, the tool modifies Docker image in Condition 1 and
exectutes the pipeline again through it, which turns to produce results
including temporary and overwritten files
(Figure~\ref{fig:overview-capturing} - (3,4)). In the second step, the
processes are labeled from the differences found in their outputs in the
two tested conditions (Figure~\ref{fig:overview-labelling}). To this order,
the tool builds docker image for Condition 2 by modifying pipeline
processes (Figure~\ref{fig:overview-labelling} - (1)) that recognised from 
the process graph. Then, the modified docker
image is executed in Condition 2 and each process is labeled right after
the running by comparing the results obtained in Condition 1
and output files of the process in Condition 2
(Figure~\ref{fig:overview-labelling} - (2)). We then label processes in 2
categories depending whether they
create differences (represented in red) or not (green). 
The tool replaces output file with the same file obtained in Condition 1 
if process creates differences (Figure~~\ref{fig:overview-labelling} - (3)).
Process labelling
is done through hooks which are defined during the docker image modifying
procedure, and incremental updates in Condition 2, until the end of the
pipeline execution.

@@

The label of processes that read or 
write temporary files is undetermined because these files are deleted during 
the execution. To address this issue, we replace in Condition 1 every process P that 
writes temporary files by a wrapper that first calls P and then
saves all its output files to a read-only directory. This replacement is
done by modifying the PATH environment variable to point to a directory
containing a modified version of the executable called by the process. The
modified executable will run the original executable and then save its
output files (see Listing~\ref{listing:modify-script}). 
This 
solution does not cover the temporary files that are removed by P 
itself, but this is not a problem since these files do not play any role in 
the subsequent steps of the pipeline.

\begin{listing}
  \begin{minted}[frame=single,
    framesep=3mm,
    linenos=false,
    xleftmargin=0pt,
    tabsize=4]{python}
  #!/usr/bin/env python
  import os.path as op
  import shutil
  import pipes

  # Run the original executable using the same arguments
  executed_file = sys.argv[0]
  exec_argv = op.join('original_scripts', executed_file)
  exec_argv = exec_argv + " ".join(map(pipes.quote, sys.argv[1:]))
  subprocess.Popen(exec_argv, shell=True).communicate()

@@

The variety of computing infrastructures, particularly the operating system, are 
known to influence the reproducibility of computational pipelines because of the
creation of small numerical differences. 
These differences mainly correspond to the mathematical functions implemented 
in different operating system libraries.
For instance, changing the mathematical functions like \emph{expf()} and 
\emph{cosf()} which manipulate the precision of floating-point representations, 
between \emph{glibc} libraries in different operating systems can produce 
small numerical differences.
Furthermore, a similar issue is expected for any operating system which is 
based on \emph{glibc}, the GNU C library.

Indeed, the propagation and amplification of these tiny perturbations by analysis 
may cause reproducibility issues. We focus on the differences 
created by different version of operating systems in computational pipelines. 
In particular, NURM-tool is tested on the neuroimaging applications which are 
predominantly using mathematical libraries.

@@

We used our tool to evaluate the effect of the operating system on results 
produced by the pipelines from the Human Connectome Project (\href{https://www.humanconnectome.org}{HCP}).
The HCP initiative is a significant effort that will acquire and analyze 
brain connectivity data from 1200 healthy adults.
It enables the neuroscience 
research community to discover relationships between brain circuits and 
individual behaviors. This helps to understand a wide range of brain disorders.
The HCP project provides (1) database services (ConnectomeDB) for storing and 
sharing primary and processed data freely, and (2) data analysis pipelines that 
are available under an open source license.

The HCP developed a set of pre-processing pipelines to process structural,
functional and diffusion MRI data acquired in the project. We focus on HCP
pre-processing pipelines for structural data, particularly PreFreeSurfer
and FreeSurfer.

According to~\cite{glasser2013}, main steps of the HCP PreFreeSurfer pipeline are:
(1) Distortion Correction: 
is the correction of geometric distortions for routine MRI scans 
to produce an undistorted native structural volume space for each subject.
(2) Anatomical Average: average same modality images (if more than one is available)
(3) ACPC (Anterior Commissure, Posterior Commissure) Alignment: 
is aligning images to the MNI space to have the same orientation as 
the reference template.
(4) Brain Extraction: includes a robust initial brain extraction, using FSL's linear (FLIRT) 
and non-linear (FNIRT) registration. 
(5) Bias Field Correction: applies intensity inhomogenity correction on the registered images.
Calculates bias field using square root of the product of T1w and T2w images. 
(6) Atlas-Registration: register the subject's native structural volume space to MNI space as the last step.

Also, the main goals of HCP FreeSurfer pipeline are:
(1) improve the robustness of brain extraction,
(2) fine tune image registration,
(3) accurately place the surface images with high resolution data,
(4) perform FreeSurfer's standard folding-based surface registration 
to their surface atlas.

A detailed description of the analyses done in the
pipelines is available in~\cite{glasser2013}.

HCP data consists of brain images collected from an individual healthy person, 
in the age range from 22-35 years~\cite{van2013wu}. Data from 1200 subjects 
is available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
Such a directory containing the HCP data is denoted by the term ``subject".
In this paper, we randomly selected 100 unprocessed subjects from 
the HCP data release S500. All the subjects were scanned using the 3-tesla 
type of MRI scanners.

\subsection{Processing of data}

We executed the pipelines using Docker containers to simplify the 
deployment of different operating system versions on execution 
platforms. 
The Docker images were built for the HCP pre-processing 
pipelines v3.19.0 (PreFreeSurfer and FreeSurfer) in 
CentOS 6.9 (Final) and CentOS 7.4 (Core). 
CentOS is a widely used operating system among the Neuroscience community because of 
the appropriate community support, network functions, safety, portability, and openness
~\cite{hanke2011neuroscience}.
These containers were installed with all the necessary software and libraries 
required to run the HCP pipelines including FSL (version 5.0.6), 
FreeSurfer (version 5.3.0-HCP, CentOS4 build), and Connectome Workbench (version 1.0).
Docker images are available on 
\href{https://hub.docker.com/r/bigdatalabteam/hcp-prefreesurfer/}{DockerHub}
for reuse.

The data were processed in the order PreFreeSurfer and FreeSurfer. 
Unprocessed subject data used as the input data for PreFreeSurfer and its output 
results used as the input data for FreeSurfer analysis in each operating system separately.
Through all the subjects, first, two runs were made on the same operating systems 
to be sure that identified differences are not coming from the operations with randomness.
Then we identified the processes in the pipelines that are responsible for 
the creation of differences using the NURM-tool.

It should be noted that some files may have the same data but 
different metadata in headers, including creation time, modification time and date, 
which turns to non-identical checksums. 
So, we added some other functions for the file comparison to be sure that differences 
are in part of the file data. 
Depending on the file type, we used functions from FreeSurfer software package to determine 
whether two files differ, including  
\emph{mri\_diff}, \emph{mris\_diff}, and \emph{lta\_diff} for comparing volumes, surface-based files, 
and lta files (transforms) respectively. Also, text files are compared using a custom function.

@@

Two types of differences can occur in the subjects due to the 
differences in the operating systems. One is inter-OS differences caused by the 
operating system library updates and the other type, intra-OS differences 
occur as a result of the pseudo-random processes used in the 
pipelines.

In our experiments, we identified the processes that create differences due to 
using random variables in the FreeSurfer pipeline. 
For example, process \emph{mris\_curvature}
computes mean and Gaussian curvatures of a cortical surface using the random variable.
To avoid such randomness we 
initialized random number generators using a seed state in 
the command line in which pipeline script calls the program.
This ensures that results are the same from run to run.
Therefore, NURM-tool can identify the process in the pipelines 
that create differences caused by the inter-OS differences.

@@

In a previous study~\cite{Scaria2017}, we showed that the HCP pre-processing 
pipelines~\cite{glasser2013} were 
sensitive to operating system variation.

Figure~\ref{fig:fnirt_result} shows differences in the \emph{T2w\_acpc\_to\_MNI\_nonli.nii.gz} files. 
These files are created as a part of ACPC-Alignment (Step 3 of PreFreeSurfer).
The RMSE (root mean squared error) value of results is 0.029 and the dice coefficient is 0.125.
The square pixels in the brain in this checkerboard images show the differences between results in two OSes. 
In the next subsection, we will show that such misalignment is created due to 
non-linear registration in \emph{FNIRT} process in PreFreeSurfer pipeline.

Furthermore, Figure 
\ref{fig:tissue_class} illustrates the binarized differences in the \emph{aseg.hires.mgz} files.
This is the result of brain tissue segmentation across OSes in FreeSurfer pipeline. 
Each color represents a different structure in the brain.
The bright spots show relevant pixels in the image which are different.
The RMSE and the dice coefficient values of results are 0.009 and 0.991 respectively.

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/segmentation.png} 
    \caption{Differences between FNIRT results from PreFreeSurfer 
    (checkerboard image \emph{T2w\_acpc\_to\_MNI\_nonli.nii.gz}) (CentOS6 vs. 
    CentOS7). Right images show the zoomed version part of the images marked 
    by the red rectangle.} 
    \label{fig:fnirt_result}
\end{figure}

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_classification.png} 
  \caption{Binarized differences between brain segmentation results from 
           FreeSurfer (CentOS6 vs. CentOS7).} 
  \label{fig:tissue_class}
\end{figure}

@@

First, subjects are clustered into different groups using the clustering method as described before.
We choosed the threshold value of zero for the clustering method. 
This turns to have the subjects with the same topology in each cluster.
Also, we used the nearest neighbor algorithm to calculate the distance between clusters.
Table~\ref{table:data-clusters} shows the PreFreeSurfer result of subject clustering. 
Among our dataset, we identified 4 types of subjects with 
different topology of \emph{process trees}. 
Differences correspond to the data processing that leads to producing disparate output file(s).
For example, subject type 3 missed output file \emph{T1w2\_gdc.nii.gz} after processing 
in comparison with subject type 1.
Besides, we verified that 
the \emph{process trees} were identical for all subjects of the same type in 
different versions of CentOS.

\begin{table}
\centering
\begin{threeparttable}
\caption{Clustering of 20 input subjects.}

\begin{tabular}{@{}llllll@{}}
\toprule
Type   & Number of Subjects  & Missing File(s)                                    \\ \midrule
type1  & 9                   & None                                               \\
type2  & 7                   & [``T1w/T1w2\_gdc.nii.gz",                          \\
       &                     &  ``T2w/T2w2\_gdc.nii.gz"]                          \\
type3  & 2                   & [``T1w/T1w2\_gdc.nii.gz"]                          \\
type4  & 2                   & [``T2w/T2w2\_gdc.nii.gz"]                          \\ \bottomrule
\end{tabular}
\begin{tablenotes}
     \small
     \item *Type refers to the topology type of the \emph{process tree} of subjects.
\end{tablenotes}
\end{threeparttable}
\label{table:data-clusters}
\end{table}

Then, we determined the frequency of the processes that create differences in 
the pipeline in Figure~\ref{fig:pfs_freq}. \emph{FLIRT} is the most frequent 
process that produces differences.
Furthermore, the percentage of processes that introduce differences 
in PreFreeSurfer along with the 
pipeline elements are depicted in Figure~\ref{fig:pfs_freq} 
including linear registration with “\emph{FLIRT}” (in 
ACPC-Alignment, BrainExtraction, DistortionCorrection, 
AtlasRegistration), non-linear registration with “\emph{FNIRT}” (in 
BrainExtraction and AltasRegistration), image warping with 
“\emph{new\_invwarp}” (in BrainExtraction and AtlasRegistration). 
Also, differences were observed in image mean and standard-deviation 
computations in image interpolation with “\emph{fslmaths}” (in AnatomicalAverage). 
It can be seen that \emph{fslmaths} and \emph{FLIRT} 
are the processes that at least have one subprocess to produce differences across all the subjects.

% \begin{figure}
%   \centering
%     \includegraphics[width=\columnwidth]{images/pfs_barchart.png} \caption{Number 
%     of occurrences of processes that create differences in PreFreeSurfer 
%     among 20 subjects.} 
%     \label{fig:pfs_chart}
%   \end{figure}

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/pfs_heatmap.png}
  \caption{Colorized processes along with the pipeline elements that introduce 
           differences among 20 subjects. The range between red and green colors show the most and 
           least frequency of processes respectively. Besides, the number 
           of occurrences of each processes with differences is represented in the line below the heatmap.}
  \label{fig:pfs_freq}
\end{figure}

\subsubsection{FreeSurfer pipeline analysis}

We found that inter-OS differences occur if pre-processed data are different in FreeSurfer. 
Figure~\ref{fig:tissue_class} shows differences that coming from PreFreeSurfer analysis 
as a result of propagation of differences. 
NURM tool identified that there are no processs in the pipeline that create differences.
This is because of using static builds of FreeSurfer which statically linked libraries (embeded libraries in the pipeline).

@@

\note{Pipelines amplify small numerical differences because they are numerically 
unstable. Furthermore, math libraries evolve over time, leading to 
different numerical errors. we listed some of the irreproducibility 
causes of the pipelines as we mentioned in the previous section 
overally along with exact command arguments.

@@

The results still suffer from small perturbations literally because of 
the fact that pipeline are not numerically stable. The preferred 
solution is to detect and fix numerical instability of the pipeline 
instead of masking the problem. These processes need to be reviewed to 
understand and correct the cause of instabilities. 
}

@@

% \begin{figure*}
% \centering
%   \includegraphics[width=.9\textwidth]{images/graph}
%   \caption{A complete process graph from the PreFreeSurfer pipeline.
% Full-resolution image available at \url{https://drive.google.com/open?id=174yyn8SuVOUcK5aRVw0bagjDanLD0FLt}.}
%   \label{fig:complete-graph}
% \end{figure*}

% \begin{figure*}
% \centering
%   \includegraphics[width=.8\textwidth]{images/hclusters}
%   \caption{Different data types clustered among 100 subjects.}
%   \label{fig:subj-clusters}
% \end{figure*}
